---
title: 'Stats 100 Project'
author: "Sumer Sao, Christian Guallpa"
date: "05-10-2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
    canonical: true
---

```{r setup, include=FALSE, echo=FALSE}

library(knitr)
library(latex2exp)
library(tidyverse)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)

```

# Introduction

In the realm of professional basketball, certain players have achieved
legendary status during the postseason, captivating fans and leaving an
indelible mark on the history of the game. One such player who has
garnered considerable attention is Jimmy Butler. Renowned for his
tenacity, leadership, and ability to deliver outstanding performances
when it matters most, Butler has become synonymous with the concept of
"Playoff Jimmy Butler." However, as the mystique surrounding his playoff
prowess grows, it becomes imperative to delve deeper and objectively
examine the notion of Playoff Jimmy Butler---whether it is a mere
illusion or a tangible phenomenon with a significant impact on the
outcomes of games and series.\

The project aims to critically analyze and investigate the existence and
impact of Playoff Jimmy Butler. While fans and experts alike have long
attributed exceptional abilities to Butler during the postseason, it is
crucial to assess whether these claims are supported by empirical
evidence or merely based on anecdotal narratives. By conducting an
in-depth examination of Butler's playoff performances, we seek to
discern whether he possesses unique qualities that truly differentiate
him from other players during the high-pressure environment of playoff
basketball.\

The project will encompass an extensive review of historical playoff
data, statistical analysis, and qualitative assessments of Butler's
on-court contributions. Moreover, we will explore various factors that
may influence the perception and performance of players in the playoffs,
including leadership qualities, clutch performances, and the ability to
elevate one's game under intense pressure. Through this comprehensive
research endeavor, we aim to shed light on the veracity of Playoff Jimmy
Butler and determine the extent to which he impacts the outcomes of
playoff games and series.\

By unearthing the truth behind the myth and unraveling the enigma of
Playoff Jimmy Butler, the project endeavors to provide a deeper
understanding of the dynamics of postseason basketball and the
significance of exceptional individual performances within this context.
Ultimately, the findings will contribute to the broader discourse on the
impact of star players during the playoffs and their ability to
influence team success in the pursuit of championship glory.

# Data Collection

The data set utilized in the project were made available on
[Kaggle](https://www.kaggle.com/datasets/nathanlauga/nba-games?resource=download)
by Nathan Lauga. These data sets were specifically collected for the
purpose of analyzing NBA games data. To compile this data set, Nathan
accessed the [NBA stats website](https://www.nba.com/stats), which
generously provides comprehensive NBA data to the public via an
accessible API endpoint. For more in-depth information regarding the
data collection process, interested individuals can refer to the GitHub
repository ["nba predictor
repo"](https://github.com/Nathanlauga/nba-data-scrapping).\

The data set comprises of five distinct files, each serving a specific
purpose. Firstly, the "games.csv" file encompasses all games played from
the conclusion of the 2003 season until the end of 2022. It includes
essential details such as the date of the game, the participating teams,
and pertinent information like the number of points scored, among
others. Additionally, the "games_details.csv" file provides intricate
game-level statistics for players, encompassing a comprehensive range of
variables. This data set offers a detailed breakdown of individual
player performance within a given game. Furthermore, the "players.csv"
file contains valuable information pertaining to the players themselves,
primarily their names and possibly other relevant details. To assess the
ranking of NBA teams on a specific day, the "ranking.csv" file comes
into play. This data set includes the rankings categorized into the
Western and Eastern conferences. Lastly, the "teams.csv" file
encompasses information about all the NBA teams, providing comprehensive
details regarding their respective identities.\

Overall, the data set, carefully compiled by Nathan Lauga, provides a
rich and diverse foundation for conducting in-depth analyses and gaining
valuable insights into NBA games and related statistics.

```{r data sets, include=FALSE, echo=FALSE}

games_details = read.csv("games_details.csv")
games = read.csv("games.csv")
players = read.csv("players.csv")
ranking = read.csv("ranking.csv")
teams = read.csv("teams.csv")

```

# Method and Validation

# Literature Review

In the article "[For Heat, the legend of 'Playoff Jimmy' continues to
grow](https://apnews.com/article/heat-jimmy-butler-playoffs-f8a16b065ba7faf991e2878185c4d559)"
published in AP news and written by Tim Reynolds, the narrative revolves
around the exceptional playoff performances of Jimmy Butler for the
Miami Heat. Despite Butler's insistence that the notion of "Playoff
Jimmy" is not a real concept, his on-court play provides compelling
evidence to the contrary. Reynolds highlights Butler's recent
extraordinary performance against the Milwaukee Bucks, where he scored a
remarkable 56 points, tying the fourth-highest scoring effort in NBA
playoff history. The article showcases Butler's offensive prowess,
including his ability to excel even against top defenders like Giannis
Antetokounmpo, Brook Lopez, and Jrue Holiday. With admiration from
teammates and opponents alike, Butler's remarkable performance
solidifies his status as one of the league's most formidable closers.
The article emphasizes Butler's humility, focus on team success, and his
commitment to the ongoing playoff journey, adding to the growing legend
of "Playoff Jimmy."\

Based on the blog post provided and the ongoing conversation, the
research will delve into the analysis of data to determine the validity
of the claims surrounding "Playoff Jimmy" Butler's exceptional
performances. The article highlights Butler's extraordinary scoring
display of 56 points against the Milwaukee Bucks and his ability to
thrive in high-pressure situations. To substantiate these claims, the
project will employ statistical analysis, comparing Butler's playoff
performances to his regular-season counterparts. By examining key
metrics such as scoring efficiency, clutch performance, and impact on
team success, we aim to objectively evaluate whether Butler's playoff
performances truly set him apart and validate the notion of "Playoff
Jimmy." Nevertheless, it is crucial to acknowledge that the legendary
performance mentioned earlier cannot be incorporated into our analysis
due to the constraints of the current data set and its recording time
frame. Therefore, our focus will primarily be on examining Jimmy
Butler's statistics spanning from 2012 until the conclusion of 2022.\

In

# Exploratory Analysis

To initiate the exploratory analysis, we will commence with a
comparative examination using a simple box plot to assess the
distribution of total points in the playoffs versus the regular season.
Upon reviewing the two box plots presented below, several observations
can be made. Firstly, the medians appear to be similar, indicating a
consistent scoring tendency. However, the upper quartile in the playoffs
demonstrates a relatively higher distribution of points . Conversely, in
the regular season, Jimmy Butler had two exceptional games with scores
exceeding 50 points, evident as outliers in the regular season boxplot.
It is worth emphasizing that in the project, evaluating Butler's overall
performance will extend beyond total points alone, as other factors will
be taken into account to assess his overall impact.

```{r playoff v regular season, include=FALSE, echo=FALSE}

butler_games_details = filter(games_details, PLAYER_NAME == 'Jimmy Butler')
butler_games_details = na.omit(butler_games_details) #Due to Injury or Other Reasons

temp = merge(butler_games_details, games, by = 'GAME_ID') #Focus on only Jimmy Butler's Games
temp$GAME_DATE_EST = as.Date(temp$GAME_DATE_EST)

playoff_stats = filter(temp, (GAME_DATE_EST >= '2012-04-28' & GAME_DATE_EST <= '2012-06-21') |
                       (GAME_DATE_EST >= '2013-04-20' & GAME_DATE_EST <= '2013-05-05') |
                       (GAME_DATE_EST >= '2014-04-19' & GAME_DATE_EST <= '2014-06-15') |
                       (GAME_DATE_EST >= '2015-04-18' & GAME_DATE_EST <= '2015-06-16') |
                       (GAME_DATE_EST >= '2016-04-16' & GAME_DATE_EST <= '2016-06-19') |
                       (GAME_DATE_EST >= '2017-04-15' & GAME_DATE_EST <= '2017-06-12') |
                       (GAME_DATE_EST >= '2018-04-14' & GAME_DATE_EST <= '2018-06-08') |
                       (GAME_DATE_EST >= '2019-04-13' & GAME_DATE_EST <= '2019-06-13') |
                       (GAME_DATE_EST >= '2020-08-17' & GAME_DATE_EST <= '2020-10-11') |
                       (GAME_DATE_EST >= '2021-05-22' & GAME_DATE_EST <= '2021-07-20') |
                       (GAME_DATE_EST >= '2022-04-16' & GAME_DATE_EST <= '2022-06-16'))

regular_stats = filter(temp, (GAME_DATE_EST < '2012-04-28') | 
                       (GAME_DATE_EST > '2012-06-21' & GAME_DATE_EST < '2013-04-20') |
                       (GAME_DATE_EST > '2013-05-05' & GAME_DATE_EST < '2014-04-19') |
                       (GAME_DATE_EST > '2014-06-15' & GAME_DATE_EST < '2015-04-18') |
                       (GAME_DATE_EST > '2015-06-16' & GAME_DATE_EST < '2016-04-16') | 
                       (GAME_DATE_EST > '2016-06-19' & GAME_DATE_EST < '2017-04-15') | 
                       (GAME_DATE_EST > '2017-06-12' & GAME_DATE_EST < '2018-04-14') | 
                       (GAME_DATE_EST > '2018-06-08' & GAME_DATE_EST < '2019-04-13') | 
                       (GAME_DATE_EST > '2019-06-13' & GAME_DATE_EST < '2020-08-17') | 
                       (GAME_DATE_EST > '2020-10-11' & GAME_DATE_EST < '2021-05-22') | 
                       (GAME_DATE_EST > '2021-07-20' & GAME_DATE_EST < '2022-04-16') |
                       (GAME_DATE_EST > '2022-06-16'))

```

```{r exploratory analysis, fig.align='center', out.width="50%", include=TRUE, echo=FALSE}

playoff_points = data.frame(Period = "PlayOffs", Points = playoff_stats$PTS)
regular_points = data.frame(Period = "Regular Season", Points = regular_stats$PTS)
df = rbind(playoff_points, regular_points)

ggplot(df) + geom_boxplot(aes(x = Period, y = Points)) + labs(y = "Points", title="Distribution of Jimmy Butler's Points") + theme_bw() + theme(plot.title = element_text(hjust = 0.5))

```

# Current Progress

# Next Steps

# Appendix

The report includes various links that can be accessed for additional
information. However, for convenience, here are the direct links to the
appendix materials:\

[Nathan's Kaggle
Post:](https://www.kaggle.com/datasets/nathanlauga/nba-games?resource=download)
<https://www.kaggle.com/datasets/nathanlauga/nba-games?resource=download>\

[NBA stats website:](https://www.nba.com/stats)
<https://www.nba.com/stats>\

[GitHub Repository "nba predictor
repo":](https://github.com/Nathanlauga/nba-data-scrapping)
<https://github.com/Nathanlauga/nba-data-scrappin>\

["For Heat, the legend of 'Playoff Jimmy' continues to
grow":](https://apnews.com/article/heat-jimmy-butler-playoffs-f8a16b065ba7faf991e2878185c4d559)
<https://apnews.com/article/heat-jimmy-butler-playoffs-f8a16b065ba7faf991e2878185c4d559>\

These links provide quick access to the relevant appendix materials for
further reference and exploration.\

The code utilized in this report is
